---
title: "EDS-223 Homework Assignment #1"
author: "Isabella Segarra"
date: 10-06-2025
format: html
theme: pandoc
execute: 
  warning: false
   message: false
   include: false
---

# Exploring Environmental Injustice

### Objective
Utilize the U.S. Environmental Protection Agency's EJ Screen data to produce two maps communicating an environmental justice issue. 


### Environment Set-up 
```{r}

#| warning: false  
#| message: false


# Load relevant libraries
library(sf) # For vector data
library(stars) # For raster data 
library(tmap) # For static and interactive maps
library(here) # For importing data 
library(tidyverse) # For data cleaning
library(dplyr) # For filtering data 


# If the libraries are not installed, please install the below packages. 

# install.packages("sf")
# install.packages("stars")
# install.packages("tmap")
# install.packages("here")
# install.packages("tidyverse")
# install.packages("dplyr")

```
### Load in data
```{r}

#| warning: false  
#| message: false
#| include: false


# Read in geodatabase of EJScreen data at the Census Block Group level

ejscreen <- sf::st_read(here::here("data", "ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))

# Read in California boundaries

california_boundaries <- sf::st_read(here::here("data", "county_boundaries", "California_County_Boundaries_and_Identifiers.shp"))

# Read in California border 

california_border <- sf::st_read(here::here("data", "ca_state", "CA_State.shp"))

# Read in California cities

california_cities <- sf::st_read(here::here("data", "California_Cities-shp", "d86f1c7b-1acd-4a0a-9413-ded642940daf202042-1-425xni.o5rsz.shp"))

```

### Filter Data 
```{r}
#| warning: false  
#| message: false
#| echo: false 

# Filter to all data from California 
california <- ejscreen %>%
  dplyr::filter(ST_ABBREV == "CA") %>% 
  janitor::clean_names()


# Filter to counties in the Central Valley 

central_valley <- california %>%
  dplyr::filter(cnty_name %in% c("Merced County", 
                                 "Butte County",
                                 "Kern County",
                                 "Colusa County", 
                                 "Kings County", 
                                 "Fresno County", 
                                 "Madera County", 
                                 "Glenn County", 
                                 "Placer County", 
                                 "Sacramento County",
                                 "San Joaquin County", 
                                 "Shasta County", 
                                 "Sutter County", 
                                 "Tehama County", 
                                 "Tulare County", 
                                 "Yolo County", 
                                 "Yuba County", 
                                 "Stanislaus County"))

# Find the average values for all variables within counties
central_valley_counties <- aggregate(central_valley, by = list(central_valley$cnty_name), FUN = mean)

# Filter California_boundaries for only Central Valley counties 

central_boundaries <- california_boundaries %>%
  dplyr::filter(CDTFA_COUN %in% c("Merced County", 
                                  "Butte County", 
                                  "Kern County",
                                  "Colusa County", 
                                  "Kings County", 
                                  "Fresno County", 
                                  "Madera County", 
                                 "Glenn County", 
                                 "Placer County", 
                                 "Sacramento County", 
                                 "San Joaquin County", 
                                 "Shasta County", 
                                 "Sutter County", 
                                 "Tehama County", 
                                 "Tulare County", 
                                 "Yolo County", 
                                 "Yuba County", 
                                 "Stanislaus County"))

# Filter california_cities to specific central valley cities (Sacramento, Fresno, Bakersfield, Stockton)

central_valley_cities <- california_cities %>% 
  dplyr::filter(AREANAME %in% c("Sacramento", "Fresno", "Bakersfield", "Stockton"))
  
```


### Map 1: Mapping Central Valley Diesel Particulate Matter

In the Central Valley, diesel pollution, primarily from heavy diesel trucks traveling throughout Highway 99 and the I-5, contribute to the region's significant air pollution (EPA, 2025). I decided to investigate which counties in the Central Valley see higher concentrations of diesel pollution. From the map, southern Central Valley counties experience higher levels of pollution. 

```{r}

diesel_map <- tm_shape(california_border) + # Add border of California
  tm_borders(fill = "#c4bd8b") +
tm_shape(central_valley_counties) + # Add the central valley counties layer 
  tm_polygons(fill = "dslpm", # Fill the counties based on diesel pollution
              fill.scale = tm_scale(values = c("#ffc100","#ff7400", "#ff0000", "darkred")),
              fill.legend = tm_legend(title = 
                                        "Diesel Particulate Matter (μg/m3)", 
                                      size = 0.5)) +
  tm_borders(col = "black", lwd = 0.5 ) + # Add borders for the counties 
  tm_title("Diesel Pollution in Central Valley Counties of California", 
           size = 1, 
           fontface = "bold") +
  tm_title("Based on EPA’s AirToxScreen", size = 0.5)+
tm_shape(central_valley_cities) + # Add central valley city names 
  tm_dots(fill = "white") +
  tm_text("AREANAME",
          col = "white",
          size = 0.5, 
          xmod = 2.2,  # Shift labels to the right
    ymod = 0 ) + 
  tm_compass(position = c("right", "top")) 
  
diesel_map # View diesel map
```

### Map 2: Mapping Central Valley Ozone 

Diesel pollution is linked to increased ozone production (EPA, 2025). Increased ozone in the atmosphere impacts the health of the ecosystem and surrounding communities. Due to this linkage, I decided to map the ozone (ppb) levels in Central Valley counties. From the map, southern Central Valley counties have increased ozone levels showing a possible connection between diesel and ozone pollution. 

```{r}

ozone_map <- tm_shape(california_border) + # Add border of California
  tm_borders(fill = "#c4bd8b") +
tm_shape(central_valley_counties) + # Add the central valley counties layer 
  tm_polygons(fill = "ozone", # Fill the counties based on ozone pollution
              fill.scale = tm_scale(values = c("#C0C6CB","#7688BB", "#304FAF", "#073763")),
              fill.legend = tm_legend(title = 
                                        "Ozone (parts per billion)", 
                                      size = 0.5)) +
  tm_borders(col = "black", lwd = 0.5 ) + # Add borders for the counties 
  tm_title("Ozone Pollution in Central Valley Counties of California", 
           size = 2, 
           fontface = "bold") +
  tm_title("Ozone annual mean top 10 of daily maximum 8-hour concentration in air. EJScreen v2.2", size = 0.5) +
tm_shape(central_valley_cities) + # Add central valley city names 
  tm_dots(fill = "white") +
  tm_text("AREANAME",
          col = "white",
          size = 0.5, 
          xmod = 2.22,  # Shift labels to the right
    ymod = 0 ) + 
  tm_compass(position = c("right", "top"))

ozone_map # View ozone map

```

